### 2016-10-26 sec-model.tex
Add citation for MeTA.
Redact MeTA's name for blind reviewing in submission mode.
Finished a preliminary pass on the model section (mainly to add a section to give an intuiive introduction to the model and explain the motivation for designing the model in this way)

\section{Model}

@@

Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
E-step. The only change is in the updating equations in the M-step,
where we replace the update for $b_i(v)$ by a pair of updates: one for
$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
\end{align}
where $a_m = o_{t,k,m}$ (for notational conciseness).

>>>>

\section{A Two-Layer HMM for MOOC Log Analysis}

\subsection{Basic idea and rationale}

Our general idea is to use a probabilistic generative model to model 
the student activities as recorded in a MOOC log, which means we will 
assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
a parameterized probabilistic model. We can then estimate the 
parameter values of the probabilistic model by fitting the model to a specific
MOOC log data set. The estimated parameter values could then be treated as 
the latent ``knowledge'' discovered from the data. Because such a generative model 
attempts to fit {\em all} the data, it enables us to discover interesting patterns
that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.

HMM is a specific probabilistic generative model with a ``built-in'' state transition system
that would control the data to be generated by the model, thus it is especially 
suitable for modeling sequence data. At any moment, the HMM would be in 
one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
another state stochastically according to a transition matrix that specifies the probability of
going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
The association of such a latent sequence of state transitions with the observed symbols makes
it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the actual behavior of a student in the following sense: First, students may be in different ``task state'' (or ``behavior state'')  in the process of study. 
For example, one task state may be to learn about a topic by mostly watching lecture videos, another task state may be to work on quizzes, and yet another may be to participate in forum discussions. While in each of these different states, the student would tend to exhibit different surface ``micro'' behaviors. For example, in the lecture study state, the student would tend to have many video-watching related behaviors and occasionally forum activities, while in the quiz-taking state (in order to pass each module), the student would tend to show many quiz-related ``micro'' activities as well as asking questions or checking disussions on the forum. Note that due to the complexity of the student behavior, it is very difficult to accurately {\em prescribe} the specific surface ``micro'' behavior patterns for each state, especially without  prior knowledge about the students. For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations. 
A major motivation for using an HMM is because 1) it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main taraget to discover and characterize, and 2) it does not make any assumption about which latent task/behavior state must be associated with which observed activities or how a student would move from one state to another, but instead let our data ``tell'' us what kind of associations are most likely and what kind of transitions are most often, and which states tend to be more long-lasting for any particular set of students.

However, if we are to use an ordinary HMM to analyze our data, we would treat each observed ``micro'' activity (e.g., video watching, or forum post reading) as an output symbol, and thus the output distribution $p(x|u)$ for each discovered latent state would be a simple distribution over all kinds of observable micro activities recorded in our log data (e.g., 50\% lecture watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading, ...). While such a distribution is meaningful and can already help us intepret the corresponding latent state, it only gives us a rather superficial characterization of student behavior. Ideally, we want $p(x|u)$ to characterize the directly observable behavior in more detail to further capture the relations and dependencies of these micro activities. To this end, we would treat an {\em entire sequence} of micro activities (e.g., one session of activities) as an observed ``symbol'' from a latent state, and further model the generation of such a sequence with another Markov model where we treat each micro activity as an {\em observable} state, and model the transitions between these activity states in very much the same way as the state transitions in HMM.  Combining this ``surface'' Markov model with the ``deep'' hidden Markov model gives us a fairly general and powerful two-layer HMM (TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior states and their transitions as well as the corresponding ``surface'' activity transition patterns associated with each latent state to facilitate interpretation and analysis of the discoverd latent state patterns. Next, we present this model more formally and discuss how to estimate its parameters.

\subsection{Formal definition of TL-HMM}

@@

Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
E-step. The only change is in the updating equations in the M-step,
where we replace the update for $b_i(v)$ by a pair of updates: one for
$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
\end{align}
where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
algorithm for HMMs with sequence observations is provided as part of the
\meta/ toolkit~\citemeta{}.
